---
title: "Crash Course in Statistics"
author: "Alex F. Bokov"
date: "03/28/2016"
output: html_document
---

```{r,echo=FALSE}
options(width=300)
rawdeid <- read.delim('deid_bmi_temp.csv',head=T);
rawdeid$AGE <- rawdeid$AGE/365;
rawdeid$BMI <- as.numeric(as.character(rawdeid$BMI));
rawdeid$TEMPERATURE <- as.numeric(as.character(rawdeid$TEMPERATURE));
rawdeid$BMI <- as.numeric(as.character(rawdeid$BMI));
rawdeid$BMI_BIN <- factor(sign(scale(rawdeid$BMI,center = 28.6)),labels=c('low','mid','hi'));
levels(rawdeid$BMI_BIN) <- c('low','hi','hi');
levels(rawdeid$SEX)<-c('f','m','m');
levels(rawdeid$RACE)<-ifelse((xx<-levels(rawdeid$RACE))%in%c('other','black','white'),xx,'other')
rawdeid$Y[rawdeid$SEX=='f']<- with(subset(rawdeid,SEX=='f'),-1+BMI/2.5+0.01*TEMPERATURE+rnorm(487178));
rawdeid$Y[rawdeid$SEX!='f']<- with(subset(rawdeid,SEX!='f'),17+0.007*TEMPERATURE-.002*AGE^2+rnorm(805297-487178));
rawdeid$Y[rawdeid$RACE=='black'&rawdeid$SEX=='f'] <- with(subset(rawdeid,RACE=='black'&SEX=='f'),Y+5-BMI/2-AGE^2*.001);
rawdeid.sam <- subset(rawdeid, PATIENT_NUM %in% unique(c(sample(PATIENT_NUM,200),sample(unique(PATIENT_NUM[SEX=='f'&RACE=='black']),30))));
```

Plot Y vs BMI_BIN
```{r}
stripchart(Y~BMI_BIN,subset(rawdeid.sam,BMI<90),method = 'jitter',jitter = 0.2,col=c('#FF000010','#0000FF10'),vertical = T,pch='.',cex=8)
```

Plot Y vs SEX
```{r}
stripchart(Y~SEX,subset(rawdeid.sam,BMI<90),method = 'jitter',jitter = 0.2, col=c('#FF000010','#0000FF10'),vertical = T,pch='.',cex=8)
```

Plot Y vs SEX _and_ BMI_BIN
```{r}
stripchart(Y~SEX+BMI_BIN,subset(rawdeid.sam,BMI<90),method = 'jitter',jitter = 0.2,col=c('#FF000010','#0000FF10'),vertical = T,pch='.',cex=8)
```

Plot Y vs SEX and _unbinned_ BMI
```{r}
plot(Y~BMI,subset(rawdeid.sam,BMI<90&SEX=='m'),ylim=c(0,30),pch='.',cex=8,col="#0000FF20");
points(Y~BMI,subset(rawdeid.sam,BMI<90&SEX=='f'),pch='.',cex=8,col="#FF000020");
```

Here is the regression model.
```{r}
sexbmi <- lm(Y~SEX*BMI,subset(rawdeid.sam,BMI<90));
summary(sexbmi);
```

Plot Y vs SEX and AGE
```{r}
plot(Y~AGE,subset(rawdeid.sam,BMI<90&SEX=='m'),ylim=c(0,30),pch='.',cex=8,col="#0000FF20");
points(Y~AGE,subset(rawdeid.sam,BMI<90&SEX=='f'),pch='.',cex=8,col="#FF000020");
```

Some evidence of non-linearity, but much better than before. But do we _really_ need _all_ these terms? How do we decide which ones to keep?
```{r}
sexbmiage.aic <- step(sexbmiage,scope=list(.~1,.~.),direction = "both");
```

Not visibly worse. But there is something else to keep in mind-- these data-points are not independent! Some of them come from the same individual sampled at multiple ages! To separately account for within-indvididual and between-individual variation, we need to use the `nlme` library.
```{r}
library(nlme);
```

Now, do some minor dark magic that we won't have time to discuss.
```{r}
lmec <- lmeControl(opt='optim',maxIter=100,msMaxIter=100,niterEM=50,msMaxEval=400,nlmStepMax=200);
```

The `lme()` function is for fitting a *L*inear *M*ixed *E*ffect model. Mixed-effect means some of the effects are "fixed", like the ones we've been using up to now, and some of them are "random"-- i.e. error terms, but now there are more than one of them. But before we do that, let's see if it's worth doing. Let's fit a `gls()` model, which doesn't use random effects, and it will allow a comparison with the `lme()` model to see if it makes a difference.
```{r}
sexbmiage.gls <- gls(sexbmiage.aic$call$formula,rawdeid.sam,subset=BMI<90,na.action=na.omit,method='ML');
summary(sexbmiage.gls)$tTable;
summary(sexbmiage.aic)$coef;
```

Now let's try fitting an `lme()` model.
```{r}
sexbmiage.lme <- lme(sexbmiage.aic$call$formula,rawdeid.sam,subset=BMI<90,method='ML',na.action=na.omit,random=~1|PATIENT_NUM);
summary(sexbmiage.lme)$tTable;
```

Is the `lme()` model a significantly better fit than the fixed-effect model? At last, something that `anova()` _is_ useful for.
```{r}
anova(sexbmiage.gls,sexbmiage.lme);
```

So far we've said: each patient has a unique baseline value, but they all have the same age and BMI effect. Let's see if that's actually true.
```{r}
sexbmiage.lmeA <- update(sexbmiage.lme,random=~AGE|PATIENT_NUM,control=lmec);
sexbmiage.lmeB <- update(sexbmiage.lme,random=~BMI|PATIENT_NUM,control=lmec);

anova(sexbmiage.lme,sexbmiage.lmeA);
anova(sexbmiage.lme,sexbmiage.lmeB);
```

So, we are better off with a random `BMI` term in addition to a random baseline. Note: we have attributed some but not all of the BMI effect to individual variation. Now there is both a fixed `BMI` effect and a random `BMI` effect. It might also be worth seeing if including both both `BMI` _and_ `AGE` as random terms further improves fit. However, this would be something to run on a fast machine over a lunch-break.
```{r}
#sexbmiage.lmeAB <- update(sexbmiage.lmeA,random=~AGE+BMI|PATIENT_NUM,control=lmec);
#anova(sexbmiage.lmeA,sexbmiage.lmeAB);
#anova(sexbmiage.lmeB,sexbmiage.lmeAB);
```

Standardized residuals look better now.
```{r}
plot(sexbmiage.lmeA);
```

Not bad! But, what if I told you there was an additional variable available: `RACE`? When I simulated the `Y` variable, for Black women only, I gave it a strong inverse correlation with `BMI` and `AGE`^2. So strong that at higher values `Y` becomes negative, and at the beginning I stipulated that a negative `Y` means the patient is in danger! Do we have enough of a sample (of unique patients) to detect this?
```{r}
cbind(table(unique(rawdeid.sam[,c('PATIENT_NUM','RACE')])$RACE))
```

Let's find out.
```{r}
try(sexbmiagerace.lmeA <- update(sexbmiage.lmeA,.~.:RACE+RACE,control=lmec,subset=BMI<90));
try(sexbmiagerace.lmeB <- update(sexbmiage.lmeB,.~.:RACE+RACE,control=lmec,subset=BMI<90));
try(anova(sexbmiage.lmeA,sexbmiagerace.lmeA));
try(anova(sexbmiage.lmeB,sexbmiagerace.lmeB));
```

So, for Black women only, this hypothetical drug would be hazardous, and strongly counter-indicated. Yet, if we used too small a sample size, or one with too few Black women in it, _we would not have noticed_. If we had added `RACE` to the model and ran `step()` it would have said that `RACE` is not a significant source of variation. In other words, any of the following can result in data being misinterpreted potentially putting patients at risk:

* Excessively small sample sizes.
* Sampling bias.
* Omitting relevant variables.
* Including too many irrelevant variables.
* Not checking the residuals on what you think is an appropriate model.

<!--
Plot Y vs SEX and TEMPERATURE
```{r}
#plot(Y~TEMPERATURE,subset(rawdeid.sam,BMI<90&SEX=='m'),ylim=c(0,30),pch='.',cex=10,col="#0000FF30");
#points(Y~TEMPERATURE,subset(rawdeid.sam,BMI<90&SEX=='f'),pch='.',cex=10,col="#FF000030");
```

Plot all the data
```{r}
#plot(subset(cbind(jitter(data.matrix(rawdeid[,2:3])),rawdeid[,4:6]),BMI<90 & TEMPERATURE>91 & TEMPERATURE<104),pch='.',cex=2,col='#00000002');
```
-->
